### 2016-10-24 latex/input/conclusion.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data. We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study and other common MIL methods. We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability. More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Bag of Words methods. CAMIL is a relatively simple and easy to implement pipeline that has both shown strong results and significant potential for even further improvement. Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than Bag of Words), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.

>>>>

We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data. We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study and other common MIL methods. We would like to emphasize that, while we used a particular type 2 diabetes dataset in this paper, the methods described here are general-purpose and could be used for any metagenomic dataset. We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability. More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Bag of Words methods. CAMIL is a relatively simple and easy to implement pipeline that has both shown strong results and significant potential for even further improvement. Future work could revolve around improving individual parts of the pipeline, such as using better assembly and clustering methods, application of different multiple instance learning methods (other than Bag of Words), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.

### 2016-10-24 latex/input/experiments.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

This section provides an overview of the different methods that we compared our pipeline to. Our pipeline uses clustering, whereas the other methods do not. Instead, those methods directly compare individual reads instead of clusters. In order to do this, sequence reads were represented as counts of k-mers. k-mers are nucleotide strings of 
length k. Since there are four possible nucleotides, the number of possible k-mers is \(4^k\). 
%For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT. 
%For a string "ATACGATA", the count for the k-mers is 2 for ATA, 1 for TAC, ACG, CGA, and GAT, and 0 for everything else. 
We wrote a script to represent the reads/contigs as vectors representing the k-mer counts in the string, and these feature 
vectors were used by the other methods. We tested possible k values 
from 1 to 6; since the number of possible k-mers is 
exponential in k, higher values for k quickly become impractical in both run time and memory usage. From experimental validation, we found a k-mer value of 3 to be the most effective. Converting the reads from their string form to k-mer vectors with k=3 took 24 hours, 8 minutes, 49 seconds.

@@

We implemented two different versions of the pipeline in Python: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction. These results are denoted as ``CAMIL D-BoW" and ``CAMIL H-BoW", respectively, in the results tables and graphs. Clustering for the pipeline with UCLUST took 14 hours, 7 minutes, 44 seconds. Without the assembly step reducing the size of the data, clustering would not have been feasible.

@@

Finally, Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM. For this comparison, we averaged the results of 10 trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set. 23 patients were used because the test set used by Qin et al. contained 23 patients \cite{qin041012}. Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant. The results held for the original 23 patient test set, where CAMIL H-BoW actually achieved 100\% accuracy.

>>>>

This section provides an overview of the different methods that we compared our pipeline to. Our pipeline uses clustering, whereas the other methods do not. Instead, those methods directly compare individual reads instead of clusters. In order to do this, sequence reads were represented as counts of k-mers. %k-mers are nucleotide strings of length k. Since there are four possible nucleotides, the number of possible k-mers is \(4^k\). 
%For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT. 
%For a string "ATACGATA", the count for the k-mers is 2 for ATA, 1 for TAC, ACG, CGA, and GAT, and 0 for everything else. 
We wrote a script to represent the reads/contigs as vectors representing the k-mer counts in the string, and these feature 
vectors were used by the other methods. We tested possible k values 
from 1 to 6; since the number of possible k-mers is 
exponential in k, higher values for k quickly become impractical in both run time and memory usage. From experimental validation, we found a k-mer value of 3 to be the most effective. Converting the reads from their string form to k-mer vectors with k=3 took 24 hours, 8 minutes, 49 seconds.

@@

We implemented two different versions of the pipeline in Python: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction. These results are denoted as ``CAMIL D-BoW" and ``CAMIL H-BoW", respectively, in the results tables and graphs. Clustering for the pipeline with UCLUST took 14 hours, 7 minutes, 44 seconds. All 367 patients were clustered together in order to keep the features consistent.

@@

Finally, Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM. We use 23 patients in the test set here because that is what Qin et al. did \cite{qin041012}. We averaged the results of 10 trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set. Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant. The results held for the original 23 patient test set used by Qin et al., where CAMIL H-BoW actually achieved 100\% accuracy.

### 2016-10-24 latex/input/methods.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

Our proposed pipeline involves a number of steps, which serve a variety of purposes. For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled. The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier. We then developed a vocabulary-based feature extraction method.

An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}. Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}. The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center. Histogram-based Bag of Words (H-BoW) methods count the number of instances from each cluster there are in each bag, instead of keeping track of the closest-matching instance to that cluster. This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important. H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on some real world problems \cite{amores13}.

Our feature extraction method uses either the Histogram-based or Distance-based BoW method. We used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy. We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in subsection \ref{deriving-instances}. Aside from the patient labels, this process is entirely de novo, and does not consult any external databases. An illustration of the pipeline is shown in Figure \ref{pipeline} on page \pageref{pipeline}.

@@

For our assembly step, we used SOAPdenovo2, because it was the assembler used in the original MGWAS study \cite{qin041012} and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}. SOAPdenovo2 first constructs a type of directed graph called a \emph{de Bruijn} graph that represents the overlaps between different sequences \cite{li10}. Reads are divided into strings of length K called \emph{k-mers}; these k-mers are the nodes of the graph \cite{zerbino08}. The choice of K is up to the user, and is important for having good assembly results. The k-mer nodes in the graph have an edge between them if a read contains those k-mers in order with an overlap of K-1 nucleotides, and the direction of the edge indicates in which order the k-mers appear \cite{zerbino08}. SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating ``tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}. This step helps to reduce the overall size of the data by eliminating some reads that would not have been useful anyway. The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}.

We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51. The average insert size was set to 350, in accordance with the reported average insert size from the MGWAS study that we used data from \cite{qin041012}. The patient files needed to be assembled separately, in order to avoid assembling reads from different patients together. Conversely, all contigs need to be in one file for clustering, to avoid inconsistent cluster assignments between different patients. Thus, we combined the contigs from each assembled patient file into a single for clustering.

@@

\subsection{Feature Extraction and Classification}

We test both the Distance-based and Histogram-based feature extraction methods within the CAMIL pipeline. For each patient, we initialized a vector with length equal to the number of clusters. The value for each scalar in that patient's feature vector would then represent either the read that matched most closely to that cluster centroid (Distance-based), or the number of reads from that cluster that were present (Histogram-based). An example of this is illustrated as a part of Figure \ref{pipeline}. We implemented this feature extraction method in Python. We performed classification with a standard SVM classifier using the generated feature vectors; in this case, we used svm-light \cite{joachims08}. The choice of classifier is not very important for BoW methods \cite{amores13}.

>>>>

Our proposed pipeline involves a number of steps, which serve a variety of purposes. For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled. The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier. We then developed a vocabulary-based feature extraction method, discussed further in subsection \ref{feature-extraction}. Using the extracted feature vectors, we trained an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy. We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in subsection \ref{deriving-instances}. Aside from the patient labels, this process is entirely de novo, and does not consult any external databases. An illustration of the pipeline is shown in Figure \ref{pipeline} on page \pageref{pipeline}.

@@

For our assembly step, we used SOAPdenovo2, because it was the assembler used in the MGWAS study \cite{qin041012} that we compare our results with and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}. It should be noted that SOAPdenovo2 was not originally intended for metagenome assembly, but is often tuned for that application, as was done by us and Qin et al. \cite{qin041012}. We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51. The average insert size was set to 350, in accordance with the reported average insert size from the MGWAS study that we used data from \cite{qin041012}. The patient files needed to be assembled separately, in order to avoid assembling reads from different patients together. Conversely, all contigs need to be in one file for clustering, to avoid inconsistent cluster assignments between different patients. Thus, we combined the contigs from each assembled patient file into a single for clustering.

@@

\subsection{Feature Extraction and Classification}
\label{feature-extraction}

We used a "vocabulary-based" feature extraction method. An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}. Step (i) is covered by our assembly and clustering process, while step (iii) is covered by performing classification with a standard SVM classifier based on the extracted feature vectors. In this case, we used svm-light \cite{joachims08}. The choice of classifier is not very important for BoW methods \cite{amores13}. Below, we describe our feature selection methods for step (ii), which we implemented in Python, as well as the rationale for using these methods.

Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}. H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on some real world problems \cite{amores13}. Thus, we tested our pipeline using both of these feature extraction methods.

Either way, the input is a set of clusters for each patient. The D-BoW method creates a feature vector based on the contig for each cluster that was the closest match to the cluster seed. For instance, say Patient A's reads include the centroid of cluster 1, another contig that has a 45\% match to the centroid of cluster 1, no contigs from cluster 2, and two contigs that match to the centroid of cluster 3, one with a 57\% match and one with an 82\% match. The string match percentage is determined by UCLUST, as described in the previous subsection. Then, D-BoW would extract the feature vector [1, 0, 0.82], indicating the contigs for Patient A that match most closely to the cluster centroid for each cluster. The H-BoW method, instead of using the closest match to each cluster, counts the number of contigs for a patient that belong to each cluster. For the above example, the H-BoW method would extract the feature vector [2, 0, 2], since Patient A has 2 representatives from clusters 1 and 3, but no representatives from cluster 2. This example is illustrated in part of Figure \ref{pipeline}.

### 2016-10-24 latex/input/related.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

The Multiple Instance Learning (MIL) problem was first described by Dietterich in the context of drug activity prediction \cite{dietterich97}. 
%In this problem, we want to figure out which of a number of different molecules will bind to a target ``binding site". 
Each molecule can assume a number of different 3-dimensional shapes (conformations), so even for molecules that are known to bind to the binding site, it is not necessarily known which conformation of the molecule succeeds in binding. If even one shape of a given molecule binds to a binding site, it is considered a ``good" molecule \cite{dietterich97}. Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains only negative instances. This is commonly referred to as the ``standard multiple instance assumption" \cite{amores13}. In the original paper, Dietterich developed a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}. In the late 1990s and early 2000s, a number of different approaches were developed for the original MIL problem, such as Diverse Density (DD) \cite{perez98}, EM-DD \cite{zhang01}, MI-SVM \cite{andrews02}, sbMIL \cite{bunescu07}, and MILES \cite{wang06}. A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}. Most of these classic methods follow the standard assumption, which is not useful in real domains in which negative bags may contain some proportion of positive instances.
%

@@

The \emph{assembly} problem involves combining overlapping short reads (usually less than 1000 base pairs) into longer sequences called \emph{contigs} (often tens of thousands of base pairs). For instance, if one read ends with the same relatively large nucleotide string that another read starts with, the reads are likely to be overlapping fragments from the same genome, and can thus be combined into one contig. This can be done either \emph{de novo} (in an unsupervised manner) or by referencing sequences against known contigs. We focus on de novo assembly, because we wish to keep our pipeline as unsupervised as possible.

One of the main purposes of assembly is to determine the whole genomes of microbial species, the vast majority of which have not or cannot be laboratory cultured, from sequencing reads \cite{zerbino08}. Even if complete genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classification, because the contigs will contain more phylogenetic and functional information than short reads. This is because short reads of less than 1000 base pairs constitute only a tiny fraction of microbial genomes, which are usually hundreds of thousands to millions of base pairs, making it difficult to ascertain much about the phylogeny of individual reads. Many modern sequence reads are produced by Next-Generation and High-Throughput Sequencers, which usually produce these short reads. Metagenomics additionally poses its own set of challenges, due to very large datasets and lack of knowledge about how many species are present and in what relative abundances \cite{namiki12}. Thus, metagenome assembly is a relatively new and challenging field. Some of the popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
%There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.

@@

The \emph{clustering} problem in this context  involves grouping input short sequences (reads or contigs)   such that sequences 
within a group are similar to each other. The clusters 
obtained from this process are referred to 
as Operational Taxonomic Units (OTUs). OTUs represent a 
group of equivalent or similar organisms. Accordingly, the 
number of OTUs in a sample gives an approximation of the 
species diversity in that sample  \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
In addition to approximating species diversity, clustering has several other key advantages. Because 
clustering is always de novo (unsupervised), it 
is not limited by the species that are 
covered in taxonomic databases. This is  important 
because it is believed that most micro-organisms that reside 
in the human body have not been laboratory cultured \cite{handelsman04}. Clustering also reduces computational 
costs by allowing analyses to operate on entire clusters instead of 
on each read/contig. Finally, clustering helps 
the classification process by allowing feature vectors to be built
at the OTU level, instead of using individual short reads. UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a}, and 
MC-LSH \cite{bibm2012} are some of the popular sequence clustering approaches.

>>>>

The Multiple Instance Learning (MIL) problem was first described by Dietterich in the context of drug activity prediction \cite{dietterich97}. 
%In this problem, we want to figure out which of a number of different molecules will bind to a target ``binding site". 
Each molecule can assume a number of different 3-dimensional shapes (conformations), so even for molecules that are known to bind to the binding site, it is not necessarily known which conformation of the molecule succeeds in binding. If even one shape of a given molecule binds to a binding site, it is considered a ``good" molecule \cite{dietterich97}. Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains only negative instances. This is commonly referred to as the ``standard multiple instance assumption" \cite{amores13}. In the original paper, Dietterich developed a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}. In the late 1990s and early 2000s, a number of different approaches were developed for the original MIL problem, such as Diverse Density (DD) \cite{perez98}, EM-DD \cite{zhang01}, MI-SVM \cite{andrews02}, sbMIL \cite{bunescu07}, and MILES \cite{wang06}. A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}. Most of these methods follow the standard assumption, which is not useful in real domains in which negative bags may contain some proportion of positive instances.
%

@@

The \emph{assembly} problem involves combining overlapping short reads (usually less than 1000 base pairs) into longer sequences called \emph{contigs} (often tens of thousands of base pairs). For instance, if one read ends with the same relatively large nucleotide string that another read starts with, the reads are likely to be overlapping fragments from the same genome, and can thus be combined into one contig. This can be done either \emph{de novo} (in an unsupervised manner) or by referencing sequences against known contigs. We focus on de novo assembly, in order to keep our pipeline as unsupervised as possible.

One of the main purposes of assembly is to determine the whole genomes of microbial species, the vast majority of which have not or cannot be laboratory cultured, from sequencing reads \cite{zerbino08}. Even if complete genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classification, because the contigs will contain more phylogenetic and functional information than short reads. This is because short reads of less than 1000 base pairs constitute only a tiny fraction of microbial genomes, which are usually hundreds of thousands to millions of base pairs, making it difficult to ascertain much about the phylogeny of individual reads. Many modern sequence reads are produced by Next-Generation and High-Throughput Sequencers, which usually produce these short reads. Metagenomics additionally poses its own set of challenges, due to very large datasets and lack of knowledge about how many species are present and in what relative abundances \cite{namiki12}. Thus, metagenome assembly is a relatively new and challenging field. Some popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}. Here we use SOAPdenovo2 because it was the assembler used by Qin et al. in their metagenomics study \cite{qin041012}.
%There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.

@@

The \emph{clustering} problem in this context  involves grouping input short sequences (reads or contigs)   such that sequences 
within a group are similar to each other. The clusters 
obtained from this process are referred to 
as Operational Taxonomic Units (OTUs). OTUs represent a 
group of equivalent or similar organisms. Accordingly, the 
number of OTUs in a sample gives an approximation of the 
species diversity in that sample  \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
In addition to approximating species diversity, clustering has several other key advantages. Because 
clustering is always de novo (unsupervised), it 
is not limited by the species that are 
covered in taxonomic databases. This is  important 
because it is believed that most micro-organisms that reside 
in the human body have not been laboratory cultured \cite{handelsman04}. Clustering also reduces computational 
costs by allowing analyses to operate on entire clusters instead of 
on each read/contig. Finally, clustering helps 
the classification process by allowing feature vectors to be built
at the OTU level, instead of using individual short reads. UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, and MC-MinH \cite{sdm2013a} are some of the popular sequence clustering approaches.
